Error Reporting Parsers: a Monad Transformer Approach

### 2013-05-18 article.tex
Jay is an author too
oops, remove double-@ in emails
extend intro parser:  nested () example, and also add explanation of it to article, and re-work explanation

\title{Error Reporting Parsers:  a Monad Transformer Approach}
\author{Matt Fenwick\email{mfenwick100@@gmail.com}}

@@

\section{Parser Combinators}
Parser combinators are an excellent approach for building expressive parsers 
that are easy to maintain.  In addition, they benefit from host-language
integration, allowing them to snarf features such as type systems and test 
frameworks.

There are many excellent papers on parser combinators, such
as this \cite{wadler} classic Wadler paper on non-deterministic parser combinators, 
Hutton's paper which covers factoring parsers into smaller pieces \cite{hutton}, 
and Leijen's Parsec paper \cite{leijen}.  
The inspiration and and much of the names used in this article are drawn 
from those references.

Let's make a quick pass through a minimal parser combinator library so that we
can see them in action before we move on to the hard stuff.  There are three
main pieces of the code:
\begin{itemize}
 \item a Parser datatype
 \item parser-specific combinators
 \item type class instances
\end{itemize}

@@

*Main> getParser (literal 'a' <|> literal 'b') "cdeab"
Nothing
\end{verbatim}
In the rest of this article, we'll be reworking these basic parser combinators,
by improving the implementations and extending the result types,
all while taking advantage of monad transformers to keep it modular.

>>>>

\title{Error Reporting Parsers:  a Monad Transformer Approach}
\author{Matt Fenwick\email{mfenwick100@gmail.com}}
\author{Jay Vyas\email{jayunit100@gmail.com}}

@@

\section{Parser Combinators}
Parser combinators are an excellent approach for building expressive, 
declarative, and easy-to-maintain parsers.  They also benefit from host-language
integration, allowing them to snarf features such as type systems and test 
frameworks.

There are many excellent papers on parser combinators, such
as Wadler's classic paper \cite{wadler} on non-deterministic parser combinators, 
Hutton's paper which covers factoring parsers into smaller pieces \cite{hutton}, 
and Leijen's Parsec paper \cite{leijen}.  
The inspiration and and much of the names used in this article are drawn 
from those references.

To get a feel for what parser combinators are and how they're used to build
parsers, let's make a quick pass through a minimal parser combinator library 
so that we can see them in action before we move on to the hard stuff.  We'll 
divide the library into three main pieces of the code:
\begin{itemize}
 \item a Parser datatype
 \item parser-specific combinators
 \item type class instances
\end{itemize}

@@

*Main> getParser (literal 'a' <|> literal 'b') "cdeab"
Nothing
\end{verbatim}
Finally, let's parse arbitrarily deeply nested parentheses.  This example 
demonstrates the use of \verb+fmap+ to apply a function to a parse result, 
\verb+check+ to make sure that \verb+char+ doesn't consume parentheses,
\verb+literal+ to match specific tokens exactly, \verb+many+ to introduce
repetition, \verb+<|>+ to express choice, and \verb+*>+ 
and \verb+<*+ to sequence parsers:
\begin{verbatim}
data Nesting
    = One Char
    | Many [Nesting]
  deriving (Show, Eq)

char :: Parser Char Nesting
char = fmap One $ check (not . flip elem "()") item

level :: Parser Char Nesting
level = literal '(' *> (fmap Many $ many element) <* literal ')'

element :: Parser Char Nesting
element = char <|> level

parseNest :: Parser Char [Nesting]
parseNest = many element
\end{verbatim}
Examples:
\begin{verbatim}
*Main> getParser parseNest "(((()))))"
Just (")",[Many [Many [Many [Many []]]]])

*Main> getParser parseNest "(()abc(def)())zy"
Just ("",[Many [Many [],One 'a',One 'b',One 'c',
                Many [One 'd',One 'e',One 'f'],Many []],
          One 'z',
          One 'y'])

*Main> getParser parseNest "(()abc(def)()"
Just ("(()abc(def)()",[])
\end{verbatim}
In the rest of this article, we'll be reworking these basic parser combinators,
by improving the implementations and extending the result types,
all while taking advantage of monad transformers to keep it modular.
